### 2025-03-14 paper/quarto_ecml/paper.tex
Parse comments from the first document
No TODOs

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations. Such explanations should mitigate a scenario in
which practitioners deploy opaque models and blindly rely on their
predictions. On countless occasions, this has happened in practice and
caused real harms to people who were adversely and unfairly affected by
automated decision-making (ADM) systems involving opaque models
\cite{oneil2016weapons,mcgregor2021preventing}. Effective XAI tools can
aid us in monitoring models and providing recourse to individuals to
turn negative outcomes (e.g., ``loan application rejected'') into
positive ones (e.g., ``application accepted''). Our work builds upon
\textbf{counterfactual explanations} (CE) proposed by
\cite{wachter2017counterfactual} as an effective approach to achieve
this goal. CEs prescribe minimal changes for factual inputs that, if
implemented, would prompt some fitted model to produce a desired output.

To our surprise, the second challenge has not yet attracted major
research interest. Specifically, there has been no concerted effort
towards improving the ``explanatory capacity'' of models, i.e., the
degree to which learned representations correspond to explanations that
are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
Def. \ref{def-explainability}). Instead, the choice has generally been
to improve the ability of XAI tools to identify the subset of
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are also compatible
with plausible explanations \cite{altmeyer2024faithful}. Fortunately,
recent findings indicate that improved explanatory capacity can arise as
a consequence of regularization techniques aimed at other training
objectives such as robustness, generalization, and generative capacity
\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
As further discussed in Section~\ref{sec-lit}, our work consolidates
these findings within a single objective.

\textbf{Specifically, we propose counterfactual training (CT)}: a novel
training regime that aligns learned representations with plausible
explanations respecting actionability constraints. The remainder of this
paper is structured as follows. Section~\ref{sec-lit} presents related
work, focusing on the link between adversarial examples and
counterfactual explanations. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to enforce global actionability constraints. In
Section~\ref{sec-experiments}, through extensive experiments, we
demonstrate that CT substantially improves explainability without
sacrificing predictive performance. We discuss the challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion} that CT is a promising approach towards
making opaque models more trustworthy.

@@

To the best of our knowledge, the proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and (global) actionability constraints. To make the
desiderata for our framework more concrete, we follow
\cite{augustin2020adversarial} in tying the concept of explainability to
the quality of CEs that can be generated for a given model. The authors
show that CEs---understood as minimal input perturbations that yield
some desired model prediction---tend to be more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom. As argued before, learned representations may be sensitive to
producing implausible explanations and mispredicting for worst-case
counterfactuals (i.e., AEs). Thus, by inducing models to ``unlearn''
susceptiblity to such examples, AT can effectively remove implausible
explanations from the solution space.

@@

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples. Pointing
to clear similarities in how CEs and AEs are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problems in representation learning. Formally,
AEs can be seen as the subset of CEs for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
certain conditions and derive theoretical upper bounds on distances
between them.

@@

\subsection{Beyond Robustness}\label{beyond-robustness}

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally faithful to the model in that they are consistent with
what the model has learned about the underlying data. Formally, they
rely on tools from energy-based modelling to minimize the divergence
between the distribution of counterfactuals and the conditional
posterior over inputs learned by the model. Their proposed
counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
and only if the underlying model has learned representations that align
with them. The authors find that both deep ensembles
\cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
\cite{grathwohl2020your} tend to do well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data \cite{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of relying
on a single locally optimal representations that corresponds to
semantically meaningless explanations for the data. Previous work by
\cite{schut2021generating} similarly found that generating plausible
(``interpretable'') counterfactual explanations is almost trivial for
deep ensembles that have also undergone adversarial training. The case
for JEMs is even clearer: they involve a hybrid objective that induces
both high predictive performance and generative capacity
\cite{grathwohl2020your}. This is closely related to the idea of
aligning models with plausible explanations and has inspired our
proposed CT objective, as we explain in Section~\ref{sec-method}.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit goal of aligning representations with plausible explanations
that comply with user requirements. In the context of CEs, plausibility
has broadly been defined as the degree to which counterfactuals comply
with the underlying data-generating process
\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. For AR recommendations to be actionable, they need to
not only result in plausible counterfactuals but also be attainable. A
plausible CE for a rejected 20-year-old loan applicant, for example,
might reveal that their application would have been accepted, if only
they were 20 years older. Ignoring all other features, this would comply
with the definition of plausibility if 40-year-old individuals were in
fact more credit-worthy on average than young adults. But of course this
CE does not qualify for providing actionable recourse to the applicant
since \emph{age} is not a (directly) mutable feature. CT aims to improve
model explainability by aligning models with counterfactuals that meet
both desiderata: plausibility and actionability. Formally, we define
explainability as follows:

>>>>

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations. Such explanations should mitigate a scenario in
which practitioners deploy opaque models and blindly rely on their
predictions. On countless occasions, this has happened in practice and
caused real harms to people who were adversely and unfairly affected by
automated decision-making (ADM) systems involving opaque models
\cite{oneil2016weapons,mcgregor2021preventing}. Effective XAI tools can
aid us in monitoring models and providing recourse to individuals to
turn negative outcomes (e.g., ``loan application rejected'') into
positive ones (e.g., ``application accepted''). In line with this, our
work builds upon \textbf{counterfactual explanations} (CE) proposed by
\cite{wachter2017counterfactual} as an effective approach to achieve
this goal. CEs prescribe minimal changes for factual inputs that, if
implemented, would prompt some fitted model to produce a desired output.

To our surprise, the second challenge has not yet attracted major
research interest. Specifically, there has been no concerted effort
towards improving the ``explanatory capacity'' of models, i.e., the
degree to which learned representations correspond to explanations that
are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
Def. \ref{def-explainability}). Instead, the choice has generally been
to improve the ability of XAI tools to identify the subset of
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are in fact
compatible with plausible explanations \cite{altmeyer2024faithful}.
Fortunately, recent findings indicate that improved explanatory capacity
can arise as a consequence of regularization techniques aimed at other
training objectives such as robustness, generalization, and generative
capacity
\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
As further discussed in Section~\ref{sec-lit}, our work consolidates
these findings within a single objective.

\textbf{Specifically, we introduce Counterfactual Training (CT)}: a
novel training regime explicitly geared towards improving the
explainability of models. In high-level terms, we define this concept as
as the extent to which valid explanations derived for an opaque model
are also deemed plausible with respect to the underlying data and the
global actionability constraints. To the best of our knowledge, our
framework represents the first attempt to address this challenge by
employing counterfactual explanations already in the training phase.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the link
between AEs and CEs. Then follow our two principal contributions. In
Section~\ref{sec-method}, we introduce our methodological framework and
show theoretically that it can be employed to enforce global
actionability constraints. In Section~\ref{sec-experiments}, through
extensive experiments, we demonstrate that CT substantially improves
explainability without sacrificing predictive performance. We discuss
the challenges in Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion} that CT is a promising approach towards
making opaque models more trustworthy.

@@

To make the desiderata for our framework more concrete, we follow
\cite{augustin2020adversarial} in tying the concept of explainability to
the quality of CEs that can be generated for a given model. The authors
show that CEs---understood as minimal input perturbations that yield
some desired model prediction---tend to be more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom. As argued before, learned representations may be sensitive to
producing implausible explanations and mispredicting for worst-case
counterfactuals (i.e., AEs). Thus, by inducing models to ``unlearn''
susceptiblity to such examples, adversarial training can effectively
remove implausible explanations from the solution space.

@@

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples. Pointing
to clear similarities in how CEs and AEs are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problems in representation learning. Formally,
AEs can be seen as the subset of CEs for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
certain conditions.

@@

\subsection{Aligning Representations with Plausible
Explanations}\label{aligning-representations-with-plausible-explanations}

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally faithful to the model in that they are consistent with
what the model has learned about the underlying data. Formally, they
rely on tools from energy-based modelling \cite{teh2003energy} to
minimize the divergence between the distribution of counterfactuals and
the conditional posterior over inputs learned by the model. Their
proposed counterfactual explainer, \emph{ECCCo}, yields plausible
explanations if and only if the underlying model has learned
representations that align with them. The authors find that both deep
ensembles \cite{lakshminarayanan2016simple} and joint energy-based
models (JEMs) \cite{grathwohl2020your} tend to do well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data \cite{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of relying
on a single locally optimal representations that corresponds to
semantically meaningless explanations for the data. Previous work by
\cite{schut2021generating} similarly found that generating plausible
(``interpretable'') CEs is almost trivial for deep ensembles that have
also undergone adversarial training. The case for JEMs is even clearer:
they involve a hybrid objective that induces both high predictive
performance and generative capacity \cite{grathwohl2020your}. This is
closely related to the idea of aligning models with plausible
explanations and has inspired our CT objective.

@@

In this section we propose our novel counterfactual training objective.
In CT, we combine ideas from adversarial training, counterfactual
explanations, and energy-based modelling with the explicit goal of
aligning representations with plausible explanations that comply with
user-defined actionability constraints.

In the context of counterfactual explanations, plausibility has broadly
been defined as the degree to which counterfactuals comply with the
underlying data-generating process
\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. To be actionable, AR recommendations must also be
attainable. A plausible CE for a rejected 20-year-old loan applicant,
for example, might reveal that their application would have been
accepted, if only they were 20 years older. Ignoring all other features,
this would comply with the definition of plausibility if 40-year-old
individuals were in fact more credit-worthy on average than young
adults. But of course this CE does not qualify for providing actionable
recourse to the applicant since \emph{age} is not a (directly) mutable
feature. Counterfactual training aims to improve model explainability by
aligning models with counterfactuals that meet both desiderata:
plausibility and actionability. Formally, we define explainability as
follows:

### 2025-03-14 paper/sections/intro.qmd
Parse comments from the first document
No TODOs

The first challenge has attracted an abundance of research on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations. Such explanations should mitigate a scenario in which practitioners deploy opaque models and blindly rely on their predictions. On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models [@oneil2016weapons; @mcgregor2021preventing]. Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted"). Our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual as an effective approach to achieve this goal. CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce a desired output.

To our surprise, the second challenge has not yet attracted major research interest. Specifically, there has been no concerted effort towards improving the "explanatory capacity" of models, i.e., the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans (see Def. \ref{def-explainability}). Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with plausible explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful]. As further discussed in @sec-lit, our work consolidates these findings within a single objective.

**Specifically, we propose counterfactual training (CT)**: a novel training regime that aligns learned representations with plausible explanations respecting actionability constraints. The remainder of this paper is structured as follows. @sec-lit presents related work, focusing on the link between adversarial examples and counterfactual explanations. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In @sec-experiments, through extensive experiments, we demonstrate that CT substantially improves explainability without sacrificing predictive performance. We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.

>>>>

The first challenge has attracted an abundance of research on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations. Such explanations should mitigate a scenario in which practitioners deploy opaque models and blindly rely on their predictions. On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models [@oneil2016weapons; @mcgregor2021preventing]. Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted"). In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual as an effective approach to achieve this goal. CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce a desired output.

To our surprise, the second challenge has not yet attracted major research interest. Specifically, there has been no concerted effort towards improving the "explanatory capacity" of models, i.e., the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans (see Def. \ref{def-explainability}). Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are in fact compatible with plausible explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful]. As further discussed in @sec-lit, our work consolidates these findings within a single objective.

**Specifically, we introduce Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explainability of models. In high-level terms, we define this concept as as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and the global actionability constraints. To the best of our knowledge, our framework represents the first attempt to address this challenge by employing counterfactual explanations already in the training phase.

 The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the link between AEs and CEs. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In @sec-experiments, through extensive experiments, we demonstrate that CT substantially improves explainability without sacrificing predictive performance. We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.

### 2025-03-14 paper/sections/method.qmd
Parse comments from the first document
No TODOs

Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit goal of aligning representations with plausible explanations that comply with user requirements. In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful]. Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models. For AR recommendations to be actionable, they need to not only result in plausible counterfactuals but also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature. CT aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability. Formally, we define explainability as follows:

>>>>

In this section we propose our novel counterfactual training objective. In CT, we combine ideas from adversarial training, counterfactual explanations, and energy-based modelling with the explicit goal of aligning representations with plausible explanations that comply with user-defined actionability constraints.

In the context of counterfactual explanations, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful]. Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models. To be actionable, AR recommendations must also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature. Counterfactual training aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability. Formally, we define explainability as follows:

### 2025-03-14 paper/sections/related.qmd
Parse comments from the first document
No TODOs

To the best of our knowledge, the proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability. In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and (global) actionability constraints. To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of CEs that can be generated for a given model. The authors show that CEs---understood as minimal input perturbations that yield some desired model prediction---tend to be more meaningful if the underlying model is more robust to adversarial examples. We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom. As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., AEs). Thus, by inducing models to "unlearn" susceptiblity to such examples, AT can effectively remove implausible explanations from the solution space.

@@

There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples. Pointing to clear similarities in how CEs and AEs are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problems in representation learning. Formally, AEs can be seen as the subset of CEs for which misclassification is achieved [@freiesleben2022intriguing]. Similarly, @pawelczyk2022exploring show that CEs and AEs are equivalent under certain conditions and derive theoretical upper bounds on distances between them.

@@

## Beyond Robustness

Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations. In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives. The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.

Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are most called for when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data. Previous work by @schut2021generating similarly found that generating plausible ("interpretable") counterfactual explanations is almost trivial for deep ensembles that have also undergone adversarial training. The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your]. This is closely related to the idea of aligning models with plausible explanations and has inspired our proposed CT objective, as we explain in @sec-method.

>>>>

To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of CEs that can be generated for a given model. The authors show that CEs---understood as minimal input perturbations that yield some desired model prediction---tend to be more meaningful if the underlying model is more robust to adversarial examples. We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom. As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., AEs). Thus, by inducing models to "unlearn" susceptiblity to such examples, adversarial training can effectively remove implausible explanations from the solution space.

@@

There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples. Pointing to clear similarities in how CEs and AEs are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problems in representation learning. Formally, AEs can be seen as the subset of CEs for which misclassification is achieved [@freiesleben2022intriguing]. Similarly, @pawelczyk2022exploring show that CEs and AEs are equivalent under certain conditions.

@@

## Aligning Representations with Plausible Explanations

Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations. In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives. The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.

Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are most called for when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data. Previous work by @schut2021generating similarly found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have also undergone adversarial training. The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your]. This is closely related to the idea of aligning models with plausible explanations and has inspired our CT objective.
